---
title: "STA130H -- Winter 2022"
author: "Your Name & Student Number"
subtitle: Module 9 Problem Set 
output:
  pdf_document: default
  html_document: default
---

```{r, echo=FALSE, include=FALSE}
library(tidyverse)
library(knitr)
library(rpart)
library(partykit)
opts_chunk$set(warning = FALSE, message = FALSE)
```

# Instructions

## How do I hand in these problems for the 11:59 a.m. ET, March deadline?

Your complete .Rmd file that you create for this problem set AND the resulting .pdf (i.e., the one you 'Knit to PDF' from your .Rmd file) must be uploaded into a Quercus assignment (link:[https://q.utoronto.ca/courses/253019/assignments/826109](https://q.utoronto.ca/courses/253019/assignments/826109)) by 11:59 a.m. ET, on March 24. Late problem sets or problems submitted outside of Quercus (e.g., by email) are *not* accepted.


# Problem set grading

There are two components to your problem set.
Part one is R based: there is 1 mandatory question in part one, and additional optional questions.

Optional R based questions are highly recommended as they will help prepare you for larger assessments (and help improve your R skills, of course!); but, they are part of the grading for the weekly problem sets.
Part two is (more) writing based: all of part two is mandatory. And this week your submission may optionally be a video or sound recording (as noted above) (and in preparation for future required video or sound recordings).

We recommend you use a word processing software like Microsoft Word to check for grammar errors in your written work. Note: there can be issues copying from Word to R Markdown so it may be easier to write in the RMarkdown file first and then copy the text to Word. Then you can make any changes flagged in Word directly in your RMarkdown file.
  
\newpage 


# Part 1

## Question 1 [REQUIRED]

Using data from the Gallup World Poll (and the World Happiness Report), we are interested in predicting which factors influence life expectancy around the world. These data are in the file `happinessdata_2017.csv`.

```{r, message=FALSE}
happiness2017 <- read_csv("happiness2017.csv")
```


### (a) Begin by creating a new variable called `life_exp_category` which takes the value "Good" for countries with a life expectancy higher than 65 years, and "Poor" otherwise. 

```{r}


```


### (b) Divide the data into training (80\%) and testing (20\%) datasets. Build a classification tree using the training data to predict which countries have `Good` vs `Poor` life expectancy, using only the `social_support` variable as a predictor.  Use the last 3 digits of your student ID number for the random seed.

```{r}
set.seed(111)  # Use last 3 digits of your student number


```

### (c) Use the same training dataset created in (b) to build a second classification tree to predict which countries have good vs poor life expectancy, using `logGDP`, `social_support`, `freedom`, and `generosity` as potential predictors.
```{r}
# Fit the tree based on training data


```

### (d) Use the testing dataset you created in (b) to calculate the confusion matrix for the trees you built in (b) and (c). Report the sensitivity (true positive rate), specificity (true negative rate) and accuracy for each of the trees.  Here you will treat "Good" life expectancy as a positive response/prediction.

```{r}
### Tree from (b)





```



```{r}
### Tree from (c)

```



### (e) Fill in the following table using the tree you constructed in part (c).  Does the fact that some of the values are missing (NA) prevent you from making predictions for the life expectancy category for these observations?

|               | `logGDP` | `social_support` | `freedom` | `generosity` | Predicted life expectancy category |
|---------------|----------|------------------|-----------|--------------|------------------------------------|
| Obs 1         | 9.68     | 0.76             | NA        | -0.35        | **ANSWER**                           |
| Obs 2         | 9.36     | NA               | 0.82      | -0.22        | **ANSWER**                           |
| Obs 3         | 10.4     | 0.88             | 0.77      | 0.11         | **ANSWER**                           |
| Obs 4         | 9.94     | 0.85             | 0.63      | 0.01         | **ANSWER**                           |



\newpage
## Question 2 [OPTIONAL]

Two classification trees were built to predict which individuals have a disease using different sets of potential predictors. We use each of these trees to predict disease status for 100 new individuals.  Below are confusion matrices corresponding to these two classification trees.

**Tree A**

|                    | Disease | No disease |
|--------------------|---------|------------|
| Predict disease    | 36      | 22         |
| Predict no disease | 2       | 40         |

**Tree B**

|                    | Disease | No disease |
|--------------------|---------|------------|
| Predict disease    | 24      | 6          |
| Predict no disease | 14      | 56         |


### (a) Calculate the accuracy, false-positive rate, and false negative rate for each classification tree. Here, a "positive" result means we predict an individual has the disease and a "negative" result means we predict they do not.



### (b) Suppose the disease is very serious if untreated.  Explain which classifier you would prefer to use.



\newpage

## Question 3 [OPTIONAL]

Data was collected on 30 cancer patients to investigate the effectiveness (Yes/No) of a treatment. Two quantitative variables, $x_i \in (0,1), i=1,2$, are considered to be important predictors of effectiveness. Suppose that the rectangles labelled as nodes in the scatterplot below represent nodes of a classification tree.  

```{r, echo=FALSE, fig.height=3, fig.width=8}
dat <- read_csv("dat.csv")
ggplot(dat, aes(x1, x2, shape = factor(type_cat), colour = factor(type_cat))) +
  geom_point(size = 2) + 
  theme_minimal() + 
  scale_color_discrete(name = "Effectiveness", breaks = c("Yes", "No")) +
  scale_shape_discrete(name = "Effectiveness", breaks = c("Yes", "No")) +
  geom_segment(aes(y = 0, yend = 0.5, x = 0.53, xend = 0.53), size = 0.1, colour = "navyblue") +
  geom_segment(aes(x = 0, xend = 1, y = 0.5, yend = 0.5), size = 0.1, colour = "navyblue") +
  geom_segment(aes(y = 0.5, yend = 1, x = 0.5, xend = 0.5), size = 0.1, colour = "navyblue") +
  annotate("text", x = 0.15, y = 0.25, label = "Node 1") +
  annotate("text", x = 0.75, y = 0.8, label = "Node 2") +
  annotate("text", x = 0.15, y = 0.8, label = "Node 3") +
  annotate("text", x = 0.75, y = 0.25, label = "Node 4")
```


### The diagram above is the geometric interpretation of a classification tree to predict drug effectiveness based on two predictors, x1 and x2. What is the predicted class of each node?  

-------------------------------------------------------------------------------------
Node    Proportion of "Yes" values in each node   Prediction (assume we declare "effective" if more than 50% of the values are "Yes")
----    ---------------------------------------   -----------------------------------
1         5/12                                       ?????

2         3/5                                        ?????

3         1/4                                        ?????

4         2/9                                        ?????
-------------------------------------------------------------------------------------


\newpage
# Part 2 [REQUIRED]

The activity this week will focus on two areas, varying your register and describing classification trees. Prior to starting the assignment, it is highly recommended that you review the infographic available here [HERE](https://q.utoronto.ca/courses/253019/pages/ps9-varying-your-register-infographic?module_item_id=3529446), and also available under Module 9.  

There are three people that you can describe classification trees to: 

- Using the 1,000 most commonly used words in the English language (see infographic for a list of the words and software to check if you used them). If you select this option, you are allowed to use words that are not part of the 1,000 most common words as long as it is a vocabulary word you are defining from the list. (e.g., you can use classification only if you are describing classification, and only when you are describing classification)..  

- A 10-year-old child. 

- A parent or guardian without a background in mathematics.

You must pick 2 of the people above to explain classification trees. You must incorporate at least 2 vocabulary words.  
 
For this assessment, you do NOT need to include an introduction or a conclusion. Rather, you should have two paragraphs, one for each scenario. The assignment cannot be more than 1 single spaced page in length. After one page the Teaching Assistant will stop reviewing your response.  


## Vocabulary 

- Classification  

- Prediction  

- Predictor(s)  

- Covariate(s)  

- Independent variable(s)  

- Dependent variable(s)  

- Input(s)  

- Output(s)  

- Training set/sample  

- Testing set/sample  

- Fitting a model  

- Confusion matrix  

- Category  

- Tree  

- Terminal node  

- Stopping rule  

- Threshold  

- True positive (sensitivity)  

- True negative (specificity)  

- False positive  

- False negative  

- Accuracy  

- Classifier  

- Node(s)  

- Terminal Node  

- Binary  
 
- Split(ting) 


## Some things to keep in mind  

- Try to not spend more than 20 minutes on your writing (plus the time to read the article).    

- Aim for more than 200 but less than 400 words.    

- Use full sentences.    

- Grammar is not the main focus of the assessment, but it is important that you communicate in a clear and professional manner (i.e., no slang or emojis should appear). 

- Be specific. A good principle when responding to a prompt in STA130 is to assume that your audience is not aware of the subject matter (or in this case has not read the prompt).  






